---
title: "Exploratory Data Analysis"
author: "Andrew Chung"
date: "2025-05-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory Data Analysis (EDA)

Andrew Chung, hc893

BTRY 4100 Final Project -- Exploratory Data Analysis.

```{r}
options(warn = -1)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(ggridges)
library(GGally)
library(patchwork)
```

## Read in Data

```{r}
station.data = read.csv("MTA_Subway_Ridership_Summarized_Apr21.csv") %>%
  filter(ridership > 1000) # apply ridership mask of 1000 (already accounted for in regression dataset)
line.data = read.csv("MTA_Subway_Line_Data_2025_Apr21.csv")

dim(station.data)
dim(line.data)
dim(model.data)
```

## EDA: Station Ridership Data

This section explores the monthly by-station ridership dataset. Each row represents a station (ridership > 1000), its ridership figure in either Month, borough, and line(s) serving it.

```{r}
head(station.data)
```

### Distribution of Ridership

First, I plotted some histograms to eyeball the distribution of ridership across stations. The raw histogram is heavily skewed right -- a logical conclusion considering the unprecedented volume of traffic seen by a select few stations (e.g. Times Square, Penn Station, Grand Central) in touristy or otherwise hectic spots in the city.

Alternatively, I also made a histogram of the log-transformed figures which appeared much more symmetric (save for a few extreme outliers on both sides) and approximately normal.

Lastly, I overlaid the density curves of ridership by month (January, February) and noticed, aside from a small left shift, a relative lack of change in station usage behavior from January to February.

```{r}
# Raw ridership
hist.1 = ggplot(station.data, aes(x = ridership)) + geom_histogram(bins = 100) + labs(
  title = "Ridership Density", x = "Ridership"
)

# Log-Transformed ridership
hist.2 = ggplot(station.data, aes(x = log(ridership))) + geom_histogram(bins = 100) + labs(
  title = "Log-Transformed Ridership Density", x = "log(Ridership)"
)

# Superimposed Densities of ridership between months
month.hist = ggplot(station.data, aes(x = log(ridership), color = as.factor(month))) + geom_density(linewidth = 1.5) + labs(
  title = "Monthly Density Curves Overlapped", x = "log(Ridership)"
)

grid.arrange(
  grid.arrange(hist.1, hist.2, nrow = 1, ncol = 2), month.hist, nrow = 2
)
```

### Station Categories

1. Frequency Table by Borough
2. Distribution of Line Counts by Station
3. Ridership Figures vs. Number of Lines Serving Station
4. Station-wise Jan vs. Feb Ridership Comparison

```{r}
# Frequency Table by Borough
freq.df <- station.data %>%
  group_by(borough) %>%
  summarise(Count = n()) %>%
  mutate(Proportion = Count / sum(Count))
borough.bars = ggplot(freq.df, aes(x = borough, y = Proportion)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Proportion of Stations by Borough",
    x = "Borough",
    y = "Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Station-Line Comparisons
station.data = station.data %>%
  mutate(split_list = strsplit(lines, ","), n_lines = lengths(split_list))
## Distribution of # of lines
nline.hist = ggplot(station.data, aes(x = n_lines, y = after_stat(density))) + geom_histogram(bins = 10) + labs(
  title = "Distribution of Line Numbers", x = "Number of Lines"
)

# Boxplot of ridership by how many lines serve the station

station.data.jan = station.data[station.data$month == 1, ]
station.data.feb = station.data[station.data$month == 2, ]

nline.box = ggplot(station.data, aes(x = factor(n_lines), y = ridership)) +
  geom_boxplot(fill = "lightcoral") +
  scale_y_log10() +
  labs(title = "Ridership vs. Number of Lines",
       x = "Number of Lines", y = "Ridership (log10)")

diff = station.data[-c(357:361, 509), ] %>%
  pivot_wider(
    id_cols = station_complex,
    names_from = month,
    values_from = ridership,
    names_prefix = "ridership_month_"
  ) %>%
  rename(jan = ridership_month_1,
         feb = ridership_month_2)

jan.feb.comp = ggplot(diff, aes(x = jan, y = feb)) +
  geom_point(alpha = 0.6) +
  #geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Jan vs. Feb Ridership per Station",
       x = "Jan Ridership (log10)", y = "Feb Ridership (log10)")

grid.arrange(
  borough.bars, 
  nline.hist, 
  nline.box,
  jan.feb.comp, 
  nrow = 2, ncol = 2
)
```

## EDA: Line Performance Data

This section explores the monthly by-line performance dataset. Each row represents a subway line, its ridership figure in either Month, and 11 distinct performance metrics rigorously compiled from 7 data sets.

```{r}
head(line.data)

# Split into January and February
line.data$month = as.numeric(format(as.POSIXct(line.data$month, format = "%Y-%m-%d"), "%m"))
line.data.jan = line.data[line.data$month == 1, ]
line.data.feb = line.data[line.data$month == 2, ]
```

### Line-Ridership Trends

1. Monthly Line Usage (Jan, Feb, Combined)
2. % Change in Monthly Line Usage from Jan to Feb

```{r}
# Ridership Bar Chart by Line (Jan, Feb, Combined)

plot.jan = ggplot(line.data.jan, aes(x = reorder(line, -num_passengers), y = num_passengers)) + geom_bar(stat = "identity") + labs(
    x = "Subway Line",
    y = "Passenger Volume",
    title = "Jan 2025" # Chart title
)
plot.feb = ggplot(line.data.feb, aes(x = reorder(line, -num_passengers), y = num_passengers)) + geom_bar(stat = "identity") + labs(
    x = "Subway Line",
    y = "Passenger Volume",
    title = "Feb 2025" # Chart title
)
plot.combined = ggplot(line.data, aes(x = reorder(line, -num_passengers), y = num_passengers)) + geom_bar(stat = "identity") + labs(
    x = "Subway Line",
    y = "Passenger Volume",
    title = "Combined" # Chart title
)
grid.arrange(plot.jan, plot.feb, plot.combined, nrow = 1)

# Percent Change in Monthly Ridership

delta_data = merge(
  line.data.jan[, c("line", "num_passengers")],
  line.data.feb[, c("line", "num_passengers")],
  by = "line"
) %>%
  rename(
    jan = num_passengers.x,
    feb = num_passengers.y
  ) %>%
  mutate(jan = jan/31, feb = feb/28) %>%
  mutate(percent_change = 100 * (feb-jan)/jan)
ggplot(delta_data, aes(x = reorder(line, -percent_change), y = percent_change)) + geom_bar(stat = "identity") + labs(
    x = "Subway Line",
    y = "Percent Change",
    title = "Percent Change of Daily Ridership" # Chart title
)
```

### Analyzing Covariate Distributions

I grouped covariates by measurement type (time (min), raw count, ratio (%)) and plotted their respective distributions as a collage of ridgeline plots.

```{r}
# Convert to a compatible format for a Ridgeline plot
line.dist.data <- line.data[, -c(1:4)] %>%
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "value"
)

# Continuous Metrics
time.plot = ggplot(line.dist.data[
    line.dist.data$feature %in% colnames(line.data)[5:6],
], aes(x = value, y = feature, fill = feature)) + geom_density_ridges() + labs(
    x = "Time (min)",
    title = "Additional Wait/Train Times" # Chart title
) + theme_minimal() + theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
)
ratio.plot = ggplot(line.dist.data[
    line.dist.data$feature %in% colnames(line.data)[7:11],
], aes(x = value, y = feature, fill = feature)) + geom_density_ridges() + labs(
    x = "Proportion",
    title = "Ratio-Based Metrics" # Chart title
) + theme_minimal() + theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
)

# Incident Counts
crit.count.plot = ggplot(line.dist.data[
    line.dist.data$feature %in% colnames(line.data)[12:13],
], aes(x = value, y = feature, fill = feature)) + geom_density_ridges() + labs(
    x = "Count",
    title = "Critical Incidents" # Chart title
) + theme_minimal() + theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
)
noncrit.count.plot = ggplot(line.dist.data[
    line.dist.data$feature %in% colnames(line.data)[14:15],
], aes(x = value, y = feature, fill = feature)) + geom_density_ridges() + labs(
    x = "Count",
    title = "Non-Critical Incidents" # Chart title
) + theme_minimal() + theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
)

# Combined Counts
line.data = line.data %>%
  mutate(infra = infra_critical + infra_noncritical,
         noninfra = noninfra_critical + noninfra_noncritical)

comb.count.plot = ggplot(line.dist.data[
    line.dist.data$feature %in% colnames(line.data)[16:17],
], aes(x = value, y = feature, fill = feature)) + geom_density_ridges() + labs(
    x = "Count",
    title = "Incidents Count" # Chart title
) + theme_minimal() + theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
)

composite.fig = grid.arrange(
  time.plot, 
  ratio.plot, 
  comb.count.plot,
  nrow = 1
) # imported to final report but not displayed on the document due to figure margin constraints.
```

### Monthly Shift in Trends

Side-by-Side Box Plots of Monthly Shift of Response and All Predictors.

```{r}
par(mfrow = c(3,4));
for (cov in colnames(line.data)[-c(1:3, 12:15)]){
  boxplot(line.data[, cov] ~ line.data$month, main = cov, xlab = "Month", ylab = NA)
} # in report
```

### Incident Counts

Tracks Count of Incidents across Lines, highlighting accident-prone services.

```{r}
plot1 <- ggplot(line.data, aes(x = reorder(line, -infra_critical), y = infra_critical)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Line",
    y = "Count",
    title = "Critical Infrastructure"
  )

plot2 <- ggplot(line.data, aes(x = reorder(line, -noninfra_critical), y = noninfra_critical)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Line",
    y = "Count",
    title = "Critical Non-Infrastructure"
  )

plot3 <- ggplot(subset(line.data, infra_noncritical > 0), aes(x = reorder(line, -infra_noncritical), y = infra_noncritical)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Line",
    y = "Count",
    title = "Non-Critical Infrastructure"
  )

plot4 <- ggplot(subset(line.data, noninfra_noncritical > 0), aes(x = reorder(line, -noninfra_noncritical), y = noninfra_noncritical)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Line",
    y = "Count",
    title = "Non-Critical Non-Infrastructure"
  )

grid.arrange(plot1, plot2, plot3, plot4, nrow = 2, ncol = 2)

line.data$infra = line.data$infra_critical + line.data$infra_noncritical
line.data$noninfra = line.data$noninfra_critical + line.data$noninfra_noncritical

ggplot(line.data, aes(x = reorder(line, -infra), y = infra)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Line",
    y = "Count",
    title = "Critical Infrastructure"
  )

ggplot(line.data, aes(x = reorder(line, -noninfra), y = noninfra)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Line",
    y = "Count",
    title = "Critical Non-Infrastructure"
  )
```

### Collinearity

```{r}
as.data.frame(cor(line.data[, 5:11]))
corr.plot = ggpairs(line.data[, 5:11])# + scale_x_continuous(n.breaks = 3) + scale_x_continuous(n.breaks = 3)
# imported to final report but not displayed on the document due to figure margin constraints.
```

### Principle Component Analysis (PCA)

<TODO>

```{r}
cont.vars = line.data[, -c(1:4, 12:15)]
res.pca = prcomp(cont.vars, scale. = T)

## scree plot (PVE)

eigenvalues <- res.pca$sdev^2
proportion_variance <- eigenvalues / sum(eigenvalues)
scree_data <- data.frame(
  PC = 1:length(eigenvalues),
  Eigenvalue = eigenvalues,
  Proportion = proportion_variance,
  Cumulative_Proportion = cumsum(proportion_variance)
)
scree.plot = ggplot(scree_data, aes(x = PC)) +
  geom_line(aes(y = Eigenvalue, color = "Eigenvalue")) +
  geom_point(aes(y = Eigenvalue, color = "Eigenvalue")) +
  geom_line(aes(y = Proportion * max(scree_data$Eigenvalue), color = "Proportion Variance")) +
  geom_point(aes(y = Proportion * max(scree_data$Eigenvalue), color = "Proportion Variance")) +
  scale_y_continuous(
    name = "Eigenvalue",
    sec.axis = sec_axis(~ . / max(scree_data$Eigenvalue), name = "Proportion of Variance")
  ) +
  scale_color_manual(values = c("Eigenvalue" = "blue", "Proportion Variance" = "red")) +
  labs(
    title = "Scree Plot for PVE",
    x = "Principal Component",
    color = "Metric"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 1:length(eigenvalues))

## Loadings (k = 3)

loadings <- res.pca$rotation

loadings.df <- as.data.frame(loadings) %>%
  mutate(variable = rownames(loadings)) %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "principal_component",
    values_to = "loading"
  ) %>%
  mutate(principal_component = factor(principal_component, levels = paste0("PC", 1:ncol(loadings))))
loadings.df

bar.plot <- ggplot(
    loadings.df %>% filter(principal_component %in% c("PC1", "PC2", "PC3")), 
    aes(x = variable, y = loading, fill = loading)
  ) +
  geom_bar(stat = "identity") +
  facet_wrap(~ principal_component, nrow = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  labs(title = "Loadings for PC1, PC2, PC3",
       x = "Covariates",
       y = "Loading",
       fill = "Loading") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )

grid.arrange(scree.plot, bar.plot, nrow = 1)
```
